### 2020-08-04 thesis/thesis.tex
[thesis] You know, more content

%%%%% WLASNE DODATKOWE PAKIETY
% \usepackage{graphicx,listings,tikz}
\usepackage{syntax}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{mathalfa}
\usepackage{csquotes}
\usepackage{textcomp}
\usepackage{stmaryrd}
\usepackage{tikz-cd}
% Not sure if necessary
\usepackage[utf8]{inputenc}
\usepackage{rotchiffre}

@@

\chapter{Introduction}

@@

TODO This requires a rework

In this thesis we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
with the addition of a few basic constructs.

\section{Problem Analysis}

TODO add citations

Algebraic effects, while not being a new concept, in the recent years
have received a lot of attention, both from theoretical and practical side.

TODO refactor the following

They give developer necessary tools for declarative programming, writing
programs in a way to strive from exposing implementation details, while still
preserving a lot of flexibility and having strong theoretical background and
ways of formal reasoning.

@@

TODO Think about merging Problem Analysis with Statement \\

@@

TODO actually point out which sections you talk about

@@

TODO Guide reader through thesis, probably later than sooner \\
TODO Clearly divide categorical from operational approach for less algebraically fluent readers

TODO it's outdated
We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with the
language is provided, as well as implementation details and examples. We conclude by
stating what are the possible augmentations, that are intended to be made in the future.

\chapter{Background}

@@

Other branch of languages theory, that we shall investigate more in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed for various use cases, most notably, type systems.

What is the source of complexity? Thanks to strong and static type systems along
with their implementations, we have solid tools to work efficiently on functions
that are pure. That being said, we claim that the core complexity of programs
comes from side effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them, can be done through monads~\cite{moggi}. However, they were
found to be, to say the least, cumbersome to work with when the number of
different effects increases. It is perhaps not a coincidence that many
functional programming languages do not have them, because of the non-composable
nature of them, or at least not composable in their implementations.

@@

where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
What in case of adding one more effect?

It turns out, that it becomes complex and unpleasent to combine two functors
together to form a new monad, for which purpose, monad transformers arose in Haskell.
Most of the common languages avoid this by not expressing computational effects
in the type system, and instead one may think about functions as being implicitly
embed in a Kleisli Category over a functor T, where T is a hidden signature over
all possible side effects that occur in our program.

@@

Not only we would like to bring back effects to our type system, but also do it
in a way that is composable. This is where algebraic effects comes to the rescue.
From theoretical point of view, we need to develop equational theory about our
effects to assert correctness of our langauge as well as to have the right hammer
to reason about our programs. That is the point where we would like to introduce
to unfamiliar reader a notion of algebraic effects.

@@

This is really a breakthrough from practical point of view, as we may substitute
logic depending on execution environment. As an example, fetching for resources
can behave differently as we run tests, debug our code, or run it on production.
In the same manner, they neatly allow us to abstract over implementation details.

Patterns like these are well known in programming, for which other alternatives
arose. One can mention interfaces from object-oriented programming, which are also
a way to describe the API of a certain component. In order to abstract from
implementation details, we pass an object around which represents a certain
interface. While this is certainly better than having no abstractions at all,
we need to pass this interface into every function that is going to use it.
This practice is called dependency injection, and while it sounds like it solves
some of the issues, we end up in functions that need to carry representants
of the interfaces, and pass them in every subcall that they make. It also can't
be ommited that it's not really a substitute, as it's solving only one particular
issue that algebraic effects address.

TODO add link to paper with plenty of examples

- Exceptions
- Async/await
- Flipping coin
- Backtracking?
- Operating on file

@@

@@

    TODO Write a coinductive proof for divergent terms using coinductive proof
    principle.

@@

    This is one of the cases where interaction with external resource multiple
    times, or more specifically in case of algebraic effects, invoking resumption more
    than once, may lead to unexpected behaviour that would not be expected in standard
    control flow. This captures the excessive generality of effects, and is the
    issue that we would like to address with coalgebraic effects.

    TODO On it's own this snippet a bit dry and of course requires additional examples,
    for instance the classic one with operating on closed file and/or on
    streams of data

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C or Python. In Python, effects are implemented through generators,
    using built-in feature of sending value when doing \textit{yield} operation.
    Resumptions in handlers that are sending value, are one-shot and tail-recursive,
    therefore we do not need to handle coalgebraic part, at the cost of flexibility.

    TODO add reference to C and Python lib

@@

\chapter{Potential Solutions}

@@

@@

\chapter{(Co) Effectful Programming}  % TODO w/o space?
\section{Examples}\label{sec:examples}

@@

    The first example will be based on modelling (nondeterministic) choice
    in the program. We will make two decisions, which will affect the computation

    result:

@@

    % Outdated
    As of this day, two implementations are available, one based on the curried
    translation and Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit stacks
    from paper. More details can be found in Section~\ref{sec:implementation}.
    All commands are available within \verb!src!  directory.

@@

\chapter{Calculus of Freak language}
\section{Syntax}

@@

\chapter{Implementation}\label{sec:implementation}

@@

\chapter{Conclusion}
\section{Summary}

>>>>

%%%%% WLASNE DODATKOWE PAKIETY
% \usepackage{graphicx,listings,tikz}
\usepackage{syntax}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{mathalfa}
\usepackage{textcomp}
\usepackage{stmaryrd}
\usepackage{tikz-cd}

@@

\chapter{Introduction}\label{chapter:introduction}

@@

In this thesis we discuss issues with excessive generality of algebraic effects,
and propose a solution without coalgebraic means along with presentation of
experimental programming language Freak, which is a language with (co) algebraic
effects and (co) handlers, where implementaiton is based on Continutation Passing
Style translation.

Algebraic effects, while not being a new concept, in the recent years have
received a lot of attention~\cite{effects-bibliography}, both from theoretical
and practical side.

They give developer necessary tools for declarative programming and writing
programs in a way to strive from exposing implementation details by defining
computational effects as an API\@. In the same time, they preserve a lot of
flexibility and have strong theoretical background.

@@

@@

TODO actually point out which sections you talk about or maybe even put example
here

@@

We start with Chapter~\ref{chapter:background}, which provides a background
about effects, defines algebraic effects in categorical setting as well as point
out dual coalgebraic effects and cohandlers, finishing with showing related work
in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
to approach issue of excesive generality,
Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
examples along with usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
operational semantics and CPS translation. In next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
augmentations, that are intended to be made in the future.

\chapter{Background}\label{chapter:background}

@@

Other branch of languages theory, that we shall investigate more in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed for various use cases, most notably, type systems.Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
they were found to be, to say the least, cumbersome to work with when the number
of different effects increases. It is perhaps not a coincidence that many
functional programming languages do not have them, because of the non-composable
nature of them, or at least not composable in their implementations.

@@

where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
What in case of adding one more effect, $f''$ and $g''$?

It turns out, that it becomes complex and unpleasent to combine two functors
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers}arose in Haskell. Most of the common
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embed in a Kleisli
Category over a functor T, where T is a hidden signature over all possible side
effects that occur in our program.

@@

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects comes to the rescue.  From theoretical point of view,
we need to develop equational theory about our effects to assert correctness of
our langauge as well as to have the right hammer to reason about our programs.
That is the point where we would like to introduce to unfamiliar reader a notion
of algebraic effects.

@@

This is really an incredible feature from practical point of view, as we may substitute
logic depending on execution environment. As an example, fetching for resources
can behave differently as we run tests, debug our code, or run it on production.
In the same manner, they neatly allow us to abstract over implementation details.

Patterns like these are well known in programming, for which other alternatives
arose. One can mention interfaces from object-oriented programming, which are also
a way to describe the API of a certain component. In order to abstract from
implementation details, we pass an object around which represents a certain
interface. While this is certainly better than having no abstractions at all,
we need to pass this interface into every function that is going to use it.
This practice is called dependency injection, and while it sounds like it solves
some of the issues, we end up in functions that need to carry representants
of the interfaces, and pass them in every subcall that they make.

TODO maybe example with comparison for these two approaches could be useful here?

That being said, it's only one particular issue that algebraic effects address.
In effects approach, handler may also drop the resumption or invoke it more than
once. There are many other great sources for getting familiarized with
effects~\cite{handlers-tutorial, programming-in-eff, koka-tutorial}, so we will
omit further explainations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

    TODO cite Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors

@@

    From categorical standpoint, coinduction is formed over a final coalgebra,
    where corecursion is the mediator between any coalgbra into a final one,
    and coinductive proof principle corresponds to uniqueness of the mediator.
    Recall that coalgebraic effects arises from a particular type of a final
    coalgebra, namely, cofree coalgebra.

    TODO Write a coinductive proof for divergent terms using coinductive proof
    principle. Introdcution to bisim, page 34,35.

@@

    This is one of the cases where interaction with external resource multiple
    times, or more specifically in case of algebraic effects, invoking resumption
    more than once, may lead to unexpected behaviour that would not be expected
    in standard control flow.

    Let's consider the following example

    TODO rethink easter eg
    \begin{verbatim}
    handle
        let fh <- do Open "praise.txt" in
        let c <- do Choice () in
        if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
        close fh
    with {
        return x -> [x] |
        choose () k -> return (append (k true) (k false)) |
    }
    \end{verbatim}

    Where we open file, and based on nondeterministic choice, we write to file,
    and then close it. This piece of code looks harmless, however, as we invoke
    resumption for the second time, we are attempting to write to a closed file,
    and then close it once again. This captures the excessive generality of
    effects, and is the issue that we would like to address with coalgebraic
    effects.

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators~\cite{one-shot}, using built-in
    feature of sending value when doing \textit{yield} operation. Resumptions in
    handlers that are sending value, are one-shot and tail-recursive, therefore
    we do not need to handle coalgebraic part, at the cost of flexibility.

@@

\chapter{Potential Solutions}\label{chapter:potential-solutions}

@@

TODO cite LINEAR USAGE OF STATE

In fact, there is a bijective translation between algebraic effects and linear
type theory, and \textit{every monad embeds in a linear state monad}.

@@

\chapter{(Co) Effectful Programming}\label{chapter:co-effectful-programming} % TODO w/o space?
\section{Examples}\label{sec:examples}

@@

    The first example will be based on modelling (nondeterministic) choice
    in the program. We will make two decisions, which will affect the computation
    result:

@@

    % Outdated
    As of this day, two implementations are available, one based on the curried
    translation and Appel~\cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit stacks
    from paper. More details can be found in Section~\ref{chapter:implementation}.
    All commands are available within \verb!src!  directory.

@@

\chapter{Calculus of Freak language}\label{chapter:calculus-of-freak-language}
\section{Syntax}

@@

\chapter{Implementation}\label{chapter:implementation}

@@

\chapter{Conclusion}\label{chapter:conclusion}
\section{Summary}
